Linear Reduction: Parsing and Interpreting Code Without Trees

### 2024-11-06 Paper/paper.tex
Updated paper

\noindent We then run LLang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies.
Doing so produces the following graph in figure \ref{fig:expr}.

@@

    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
    \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,n)$}
    \label{fig:expr}

@@

>>>>

\noindent We then run LLang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies (figure \ref{fig:expr1}) and
expressions generated by $\hbox{\tencsc Generate-Expression}(n,1,100)$ as $n$ varies (figure \ref{fig:expr2}).

@@

    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
    \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,n)$}
    \label{fig:expr1}
    \medskip
    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr1.png}}
    \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,1)$}
    \label{fig:expr2}

@@

\noindent Running each implementation 5 times for each $x$, we get the following graph in figure \ref{fig:fib} (as $x$ varies):

    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
    \caption{Time to compute $\hbox{\tt fib}(x)$}
    \label{fig:fib}

Our implementation takes approximately $\sim ab^x$ seconds to run where $a=0.000138585, b=1.64685$, and the menhir-based interpreter takes $\sim ab^x$ seconds where $a=0.00000362553, b=1.5809$.

\subsection{Conclusion}

We can see that linear reduction is significantly slower than the menhir-based interpreter.
This can be for various reasons:
\begin{enumerate}
    \item Linear reduction is an $O(n^2)$-time algorithm (when no reduction produces a string of printable tokens, and every reduction reduces two tokens to one).
        This is because a string of the form $\s s^1_1\s s^2_2\cdots\s s^n_n$ will take $n$ applications of the $\beta$-reducer, each application taking $n-k$ iterations (since $\beta$ is recursive).
        This can be made more efficient: instead of starting the next $\beta$-reduction at the beginning of a string, it can be started at the token before the one just reduced.
        Meaning once $\s s^{n-1}$ and $\s s^n$ are reduced to give $\s x^{n-1}$, we can begin the next iteration of the $\beta$-reducer at $\s s^{n-2}\s x^{n-1}$.
        Unfortunately, implementing this did not lead to substantial increases in efficiency.

    \item When handling function calls, linear reduction simply re-reduces the string in the definition of the function.
        This requires it to parse and interpret the function again.
        But classically, since the function definition has already been parsed and a parse tree created, further parsing of the function is not created.
        This means that classical methods handle function calls better than linear reduction, but on the other hand linear reduction can handle macro-like programming better.
        For example, if one were to change the state to map \ttt( to $\Lbrace$, the menhir-based interpreter would not be able to reparse the function to reinterpret \ttt(.
        But this naturally happens with linear reduction, for better or worse.
        \gobble)\gobble)

    \item The standard methods of parsing code are well-studied and optimized.
        Menhir does not naively parse code, it is an optimized tool.
        Thus comparing a naive implementation of linear reduction to menhir is not a fair comparison.
\end{enumerate}

\section{Producing Parse Trees}

Linear reduction can be used to just parse, instead of parsing and interpreting simultaneously.
This has its benefits, as separating parsing and interpretation allows for the inspection of code before running it, which can help alert the programmer to syntax errors.
Furthermore, this means linear reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.

To do so, we define $\tree$, the set of trees, as follows:
\[ \tree \ccoloneqq \Sigma \gbar \Sigma(\tree) \gbar \Sigma(\tree)(\tree) \gbar \Sigma(\tree)(\tree)(\tree) \gbar \cdots \]
So a tree is simply a multi-branched tree whose nodes are elements of our alphabet $\Sigma$.
Then we define an \textit{initial $\beta$-reducer} as
\[ \hb\colon \Sigma\times\Sigma \partial \Sigma \times (\bbZ\times\bbZ\to\bbZ) \times (\tree\times\tree\to\tree) \]
This extends to a \textit{derived $\beta$-reducer}
\[ \beta\colon (\oS\times\tree)^* \longto (\oS\times\tree)^* \]
where $\beta(\xi)$ is defined as
\begin{enumerate}
    \item If $\xi=\sigma^1_n(t)\sigma^2_m(s)\xi'$ where $n\geq m$ and $\hb(\sigma^1,\sigma^2)=(\tau,\rho,f)$ is defined then $\beta(\xi)=\tau_{\rho(n,m)}(f(t,s))\xi'$.
    \item Otherwise $\beta(\xi)=\sigma^1_n(t)\beta(\sigma^2_m(s)\xi')$.
\end{enumerate}
We can also add the case that if $\xi$ is a single token then $\beta(\xi)=\xi_0$.

For example, for arithmetic we can define
\blist
    \item $\NUM\ \OP\to\OP\NUM,\snd,(t,f\mapsto f(t))$
    \item $\OP\NUM\ \NUM\to \NUM,\snd,(f(t),s\mapsto f(t)(s))$
    \item $\OP\NUM\ \OP\NUM\to \OP\NUM,\snd,\bigl(f(t),g(s)\mapsto g(f(t)(s))\bigr)$
    \item $\NUM\ \RPAREN\to \RPAREN\NUM,\snd,(t,\_\mapsto t)$
    \item $\OP\NUM\ \RPAREN\NUM\to \RPAREN\NUM,\snd,(f(t),s\mapsto f(t)(s))$
    \item $\LPAREN\ \RPAREN\NUM\to \NUM,\fst,(\_,t\mapsto t)$
\elist

So for example, \ttt{1 + 2 * (3 + 4)} is then parsed as
\begin{align*}
    \mb1_\infty \mb+_1 \mb2_\infty \mb*_2 \mb(_\infty \mb3_\infty \mb+_1 \mb4_\infty \mb)_0 &\to^* \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)}_1 \mb{4}_0\\
    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb(_\infty \mb{+(3)(4)}_0\\
    &\to \mb{+(1)}_1 \mb{*(2)}_2 \mb{+(3)(4)}_\infty\\
    &\to^* \mb{+(1)}_1 \mb{*(2)(+(3)(4))}_0\\
    &\to \mb{+(1)(*(2)(+(3)(4)))}_0
\end{align*}
\gobble)\gobble)
